\chapter{Introduction}
\label{chap:introduction}

\section{Motivation}
\label{sec:motivation}

Large language models are increasingly deployed as personal-decision
advisors, yet little is known about whether the advice they give
reflects the diversity of views that real human communities express on
the same decisions. Conversational assistants now field hundreds of
millions of weekly queries, and a growing share of those queries ask
for guidance on consequential personal choices rather than for
factual lookup or code completion \citep{chatterji-2025-chatgpt,
tamkin-2024-clio}. Education, career, and finance decisions are the
canonical example: they are high-stakes, they shape life trajectories,
and they are often genuinely contested. People ask peers, forums, and
now language models precisely because there is no single right answer.

When a community of informed humans is split roughly evenly on whether
to take a job offer or change a major, an aligned advisor should
reflect that split rather than collapse to one side. The empirical
record, however, points the other way. Recent work on moral judgment
shows that frontier models converge on a single answer where humans
disagree, even when prompts are identical and the underlying question
admits multiple defensible positions \citep{russo-2026-pluralistic}.
Whether the same collapse occurs on personal-decision advice, and
whether it can be diagnosed and reversed, is the question this thesis
takes up.

\section{The puzzle}
\label{sec:puzzle}

Across four state-of-the-art language models and 132 ambiguous Reddit
decision posts, the models gave near-deterministic answers even where
the community was split roughly evenly. When the upvote-weighted
distribution of comments was near 50/50, the models still answered
close to 100/0. The gap is not a small drift in calibration; the two
distributions have fundamentally different shapes --- one broad, one
nearly degenerate. We quantify this gap and report the per-model
entropy values in \cref{chap:direction1}.

The puzzle is mechanism. Two explanations are observationally similar
in the standard setup but have opposite deployment consequences. The
first is \emph{belief collapse}: the model's internal posterior over
options is genuinely narrow, in which case alignment requires
retraining or new preference data \citep{kirk-2024-rlhf-diversity}.
The second is \emph{decoding collapse}: the internal posterior is
broad, but a forced single-token answer interface discards everything
except the mode \citep{zhang-2025-verbalized-sampling}. The first is a
training-level problem and is expensive to fix; the second is a
deployment-level problem and is fixable with a prompt change. Standard
benchmarks elicit a forced choice and so cannot tell the two apart.

A parallel literature attributes the gap to role framing: models cast
as advisors are said to behave more prescriptively than models cast as
neutral describers \citep{russo-2026-pluralistic}. If true at scale,
this would be a third mechanism, intermediate in cost between training
and prompting. Whether output format and role framing are equally
important, or whether one dominates the other, has not been measured
on personal-decision data.

\section{What this thesis contributes}
\label{sec:contributions}

We demonstrate that LLM choice collapse on personal decisions is
format-dependent rather than format-invariant --- it is governed by the
output format the model is queried under, not by the content of the
decision --- and that role framing, though statistically detectable, is
16 times weaker than output format as a lever for pluralism. The
contributions are four.

First, we release the first personal-decision alignment dataset built
for this purpose: 811 real Reddit posts spanning education, career,
and finance, all posted after the knowledge cutoff of every model we
evaluate, with full comment-level stance distributions preserved as
the alignment target. Posting-after-cutoff eliminates training
contamination as a confound, which prior personal-advice work has not
ruled out \citep{sachdeva-2025-aita}.

Second, we provide direct mechanism evidence. Switching the output
format from forced choice to verbalized sampling, while holding the
prompt and temperature otherwise constant, recovers 94--101\% of the
entropy gap across the 4 models tested. A model that recovers
near-human entropy under a different output format demonstrates that it
can express a broad distribution when the format permits, whatever its
internal state may be \citep{zhang-2025-verbalized-sampling}.

Third, we rank the two interface mechanisms against each other.
Output format reduces the absolute alignment gap by 0.235 (Cohen's $d
\approx 1.6$); role framing reduces it by 0.015 ($d \approx 0.04$).
The ratio is roughly 16$\times$. The contrast itself is the finding:
the mechanism that the alignment-on-advice literature has emphasized
is the smaller of the two by an order of magnitude.

Fourth, the result is a deployment recommendation. A prompt-only
intervention recovers near-human pluralism on this benchmark without
any model retraining, fine-tuning, or preference data
\citep{russo-2026-pluralistic}. Whether this generalizes beyond the
personal-decision domain is left to future work; within the domain,
the practical implication is that the existing fleet of frontier
models can already express the diversity that aligned advice requires,
and the deployment surface is what discards it.

\section{Thesis statement and research questions}
\label{sec:rqs}

\begin{quote}
LLM advice on personal decisions exhibits a pluralistic gap that is
structural and format-dependent, not stimulus-driven. The gap is a
deployment-level problem, not a training-level problem, and is
recoverable through interface design. We make no claim about the
model's internal state: our evidence concerns expressed behavior under
different elicitation formats.
\end{quote}

\noindent We address this through three research questions.

\begin{enumerate}
  \item \textbf{RQ1.} To what extent do LLM choice distributions on
        personal decisions diverge from the human community choice
        distributions on the same posts, in mean and in distributional
        shape?
  \item \textbf{RQ2.} Is the divergence format-dependent (the expressed
        distribution changes when output format changes) or
        format-invariant (the expressed distribution stays peaked
        regardless of format, implying retraining is required)?
  \item \textbf{RQ3.} Is the divergence role-induced (recoverable by
        shifting from prescriptive advisor framing to descriptive
        oracle framing)?
\end{enumerate}

\section{Thesis structure}
\label{sec:structure}

\Cref{chap:related_work} reviews the literature on RLHF-induced mode
collapse, pluralistic alignment, and decoding-time diversity, locating
the gap that this thesis fills \citep{ouyang-2022-instructgpt,
bai-2022-constitutional}. \Cref{chap:methodology} describes the data
pipeline, the four-model evaluation grid, and the verbalized-sampling
protocol that operationalizes RQ2 and RQ3. \Cref{chap:direction1}
presents the Direction 1 results: forced choice versus verbalized
sampling across 132 ambiguous posts, addressing RQ1 and RQ2.
\Cref{chap:direction2} presents the Direction 2 results: advisor
versus oracle versus personal framing across 811 posts, addressing
RQ3. \Cref{chap:discussion} interprets the 16$\times$ mechanism ratio
and discusses why content-level moderators (reversibility, time
horizon, socioeconomic context) show null effects in the
mixed-effects regression. \Cref{chap:conclusion} states deployment
recommendations and the limits of generalization beyond the
personal-decision domain.
