\chapter{Related work}
\label{chap:related_work}

This thesis sits at the intersection of four lines of prior work:
deployment-side studies of LLMs as advice systems, pluralistic
alignment, RLHF-induced diversity loss and its decoding-time
remedies, and prompt-framing sensitivity.
\Cref{sec:advisors,sec:pluralistic,sec:rlhf-vs,sec:framing} review each
line and locate the specific gap that the thesis fills;
\cref{sec:reddit-baseline} discusses Reddit as a community-baseline
source; \cref{sec:synthesis}
synthesizes the four gaps and ties them to RQ1--RQ3.

\section{LLMs as decision advisors}
\label{sec:advisors}

Large language models are now widely used as advice systems for
high-stakes personal decisions, yet evaluation of their advice has
lagged behind their deployment. Usage logs show that conversational
assistants field a large and growing share of advice-seeking queries
on career, education, and personal-finance topics
\citep{chatterji-2025-chatgpt, tamkin-2024-clio}. Anthropic's Clio
analysis classified millions of Claude conversations and found that
guidance and advice tasks form a top-tier cluster, distinct from
coding and information lookup \citep{tamkin-2024-clio}. Subsequent
economic-task analyses corroborate this pattern at scale and across
labour-market categories \citep{handa-2025-economic-tasks}.

Evaluation work has begun to catch up but remains narrow. The closest
comparable benchmark is \citet{kumar-2026-when-ai-gives}, who
compared LLM and human responses on well-being posts and found
systematic differences in tone, advice content, and recommendation
diversity. \citet{cahyono-2025-life-changing} probed LLMs on
life-changing decisions but used a small synthetic stimulus set and
did not preserve a community-level human distribution. Across this
line of work the unit of comparison is typically a single
``ground-truth'' answer or a panel of expert annotators, not the
natural distribution of opinions that arises in a real online
community.

\textbf{Gap:} no existing benchmark pairs a large set of real
personal-decision posts with the full upvote-weighted comment
distribution that an aligned advisor should ideally reflect.

\section{Pluralistic alignment and the moral gap}
\label{sec:pluralistic}

\citet{russo-2026-pluralistic} introduced the concept of a
pluralistic moral gap: LLMs match human distributions on consensus
dilemmas but collapse to majority answers on ambiguous ones. Their
Moral Dilemma Distributions (MDD) dataset paired AITA-style scenarios
with fine-grained human verdict distributions, and their
distributional alignment task scored models on JS divergence rather
than top-1 accuracy. They then proposed a Distributional Moral
Probing (DMP) intervention and reported that the model--human gap
was substantially larger on ambiguous AITA dilemmas (gap $\approx 0.30$)
than on high-consensus items (gap $\approx 0.18$).

The methodological precedent for using Reddit comment label rates as
a community baseline is \citet{sachdeva-2025-aita}, who showed that
upvote-weighted YTA / NTA labels in r/AmItheAsshole produce stable,
reproducible community distributions across thousands of dilemmas
and serve as a credible target for distributional alignment. The
broader pluralistic-alignment programme generalizes the same
intuition beyond moral judgment:
\citet{sorensen-2024-value-kaleidoscope} mapped plural human values
onto a structured kaleidoscope of rights and duties, and
\citet{sorensen-2024-roadmap} laid out a research agenda for
distributional, steerable, and overton-pluralistic models.

This line of work establishes that the gap exists and that it is
measurable at the distribution level, but it stops short of two
things. The stimulus is moral judgment (AITA), which is structurally
different from open-ended decision advice: the answer space is
fixed, the verdicts are categorical, and the question being asked is
about blame rather than about what the protagonist should do. And
the mechanism behind the collapse is not decomposed --- the gap is
documented but not attributed to format-invariant expression,
format-dependent expression, or framing.

\textbf{Gap:} the pluralistic-alignment literature is bounded to
moral judgment and does not separate format-dependent from
format-invariant expression.

\section{RLHF, mode collapse, and verbalized sampling}
\label{sec:rlhf-vs}

Reinforcement learning from human feedback compresses output
diversity, producing modal outputs that are detectable across
generation tasks. Instruction-tuned models trained with RLHF
\citep{ouyang-2022-instructgpt} and constitutional AI
\citep{bai-2022-constitutional} are reliably more peaked than their
base counterparts on the same prompts.
\citet{kirk-2024-rlhf-diversity} measured this across a battery of
generation tasks and found a consistent reduction in lexical and
semantic diversity post-RLHF, and \citet{padmakumar-2024-diversity}
extended the observation to human-AI co-writing, where collaborating
with an RLHF model demonstrably narrowed the writer's own output
distribution.

A theoretical account is emerging. \citet{chung-2025-soft-preference}
argued that the combination of the KL regularizer in PPO and a
typicality bias in the preference data drives the optimum towards
modal completions, and that the loss of mass on minority but
admissible answers is a property of the objective rather than of any
specific dataset.

The most direct decoding-time remedy is verbalized sampling.
\citet{zhang-2025-verbalized-sampling} asked models to output a
distribution over candidate answers in a single forward pass --- for
example, ``produce 5 answers and their probabilities'' --- and
recovered the diversity lost to RLHF --- a 2--3$\times$ diversity gain. The technique
was validated on creative writing, joke generation, donation
amounts, and US-state enumeration, but never on advice tasks or on
forced-choice decision settings where the answer space is binary or
small and the goal is calibrated probability rather than novelty.

\textbf{Gap:} verbalized sampling has not been tested on advice or
on forced-choice decisions, leaving open whether the same
decoding-time fix transfers to settings where alignment, rather
than creativity, is the target.

\section{Sycophancy, framing, and prompt sensitivity}
\label{sec:framing}

LLM outputs depend systematically on how questions are framed,
raising the question of whether the alignment gap is also
framing-induced. \citet{cheng-2026-elephant} introduced ELEPHANT, a
benchmark that distinguishes social sycophancy --- face-preserving
agreement and emotional validation --- from propositional
sycophancy, and showed that frontier models exhibit social
sycophancy in open-ended advice even when their underlying judgments
stay stable. \citet{fanous-2025-syceval} reported that propositional
sycophancy also persists at high rates: 78.5\% of models maintained
an incorrect answer after a confident user rebuttal in their setup.

Beyond agreement, models are sensitive to surface framing of the
question itself. \citet{vannuenen-2026-fragility} found that 24.3\%
of AITA verdicts flip when the point of view of the narrator is
changed without altering the underlying facts of the dilemma, and
\citet{pezeshkpour-2024-position-bias} documented strong position
biases in multiple-choice tasks where simply reordering the answer
options changes which one the model picks.

These findings motivate the role-framing question of the present
thesis: if the alignment gap collapses under a different
output-format prompt, does it also collapse under a different role
frame? Prior work in this line is bounded to moral judgment or
synthetic MCQ benchmarks and has not measured framing effects in
personal-decision forced-choice tasks.

\textbf{Gap:} no prior work has tested advisor versus oracle versus
personal framing on personal-decision forced-choice tasks, so the
size of the framing effect relative to the output-format effect is
unknown.

\section{Reddit as a source of human community baselines}
\label{sec:reddit-baseline}

Reddit comment distributions provide an upvote-weighted
operationalization of community consensus that scales to thousands
of decisions but reflects the views of vocal commenters rather than
the general public. The methodological case for treating
upvote-weighted labels as a community signal is made by
\citet{sachdeva-2025-aita}, and the data infrastructure for working
at scale --- monthly dumps, deletion-respecting archives, and
standardized fields --- comes from the Pushshift project
\citep{baumgartner-2020-pushshift}.

The limitation is selection. Vocal commenters in r/personalfinance
skew toward FIRE-style ideology; r/cscareerquestions skews toward
high-cost-of-living tech labour markets; r/college skews toward
US-centric four-year-degree narratives. Throughout this thesis we
therefore call the baseline ``Reddit community consensus'' or
``upvote-weighted comment distribution'', never ``human consensus''
in isolation. Training contamination is a separate concern,
mitigated here by restricting the corpus to posts written after the
knowledge cutoff of every model evaluated.

A practical observation: most Reddit-LLM studies to date use
r/AmItheAsshole exclusively
\citep{russo-2026-pluralistic, sachdeva-2025-aita,
vannuenen-2026-fragility}. Personal-decision subreddits in
education, career, and finance --- where the question is what the
protagonist should \emph{do} rather than who is at \emph{fault} ---
are effectively unstudied.

\textbf{Gap:} the methodology of upvote-weighted Reddit baselines is
established for moral judgment but has not been ported to
personal-decision forums, where the answer space and the rhetorical
function of comments are different.

\section{Synthesis}
\label{sec:synthesis}

The four lines above leave four distinct gaps. The advice-evaluation
literature lacks a large real-Reddit personal-decision dataset with
preserved human distributions; the pluralistic-alignment literature
documents the gap but does not decompose its mechanism and is
bounded to moral judgment; the RLHF-diversity literature offers a
decoding-time remedy that has not been tested on advice or on
forced-choice settings; and the framing-sensitivity literature has
not measured role-frame effects in personal-decision tasks. RQ1
addresses the first gap by quantifying the divergence on a 132-post
ambiguous-decision benchmark, RQ2 addresses the second and third
gaps jointly by testing whether verbalized sampling closes the
divergence (a positive result rules out format-invariant expression),
and
RQ3 addresses the fourth gap by measuring the size of the
advisor-versus-oracle framing effect against the output-format
effect on the same posts.
