\chapter{Conclusion}
\label{chap:conclusion}

%% -----------------------------------------------------------------------
\section{Three contributions}
\label{sec:conc_contributions}

This thesis makes three contributions to the study of LLM--human
alignment on personal-decision advice.

\paragraph{Empirical contribution.}
We constructed and analysed the first cross-model alignment dataset
for personal-decision advice: 620 real Reddit submissions in
education, career, and finance, all posted after the knowledge cutoff
of every study model, paired with their upvote-weighted comment
distributions as the alignment target. Across four post-trained LLMs,
the distributional gap between LLM forced-choice outputs and the
community baseline is large and consistent: a mean entropy gap of
0.575 nats on the ambiguous subset, with a Cohen's $d$ of 3.58. The
dataset, with built-in training-contamination control through the
post-cutoff filter, is the primary empirical asset this thesis
contributes to the field.

\paragraph{Mechanistic contribution.}
Through a verbalized-sampling diagnostic on the ambiguous subset, we
demonstrated that the alignment gap is format-dependent rather than
format-invariant. A prompt-level change --- asking the LLM to output a
probability distribution over options, rather than commit to a single
forced choice --- recovered 94 to 101 percent of the entropy gap
across all four study models. The recovery is robust to a temperature
control: raising sampling temperature from 0.7 to 1.2 in the same
forced-choice format closed only a small fraction of the gap that
distributional output closed at the lower temperature. The models'
capacity to express a broad distribution over personal-decision options,
in other words, is preserved through current post-training paradigms;
what collapses at deployment is the interface, not that capacity.

\paragraph{Theoretical contribution.}
We developed the \textbf{interface-level vs training-level
distinction} as an analytic framework for organising the space of
LLM alignment interventions. Output format (single token vs.
distribution) operates at the interface level: it determines what the
model is permitted to express. Role framing (advisor vs. oracle)
operates closer to the training level: it modulates what the model is
taught to treat as appropriate. These two levers' effects on
absolute alignment gap differ by approximately 16 times --- the
empirical signature of the distinction. The framework locates existing accounts of LLM
sampling --- the typicality account of
\citet{zhang-2025-verbalized-sampling} and the descriptive--prescriptive
account of \citet{sivaprasad-2025-descriptive-prescriptive} ---
within a sharper question: at what point in the LLM stack can
pluralism be recovered, and at what cost?

%% -----------------------------------------------------------------------
\section{The work in context}
\label{sec:conc_context}

This thesis contributes one piece of evidence to a longer conversation
about whether and how language models can reflect the pluralism of
human communities.

The piece is small. One task type (binary forced choice); four
models, all post-trained; three decision domains on Reddit; one month
of post-cutoff data; one annotator pipeline. Findings here cannot
speak to multi-turn deliberation, multi-option decisions, or the
advice that LLMs give to non-English-speaking communities. The
absolute alignment numbers describe alignment with the Reddit
community baseline used here, not with humanity in general.

The mechanism account we propose may not survive scrutiny in other
tasks. It is plausible that in domains where post-training pressures
are stronger --- safety filtering, refusal behaviour, factual
recall --- the balance between interface and training levers shifts.
The distinction is meaningful for personal-decision advice; whether it
generalises is empirically open.

If the mechanism account holds up beyond this thesis, it suggests an
organising principle for LLM alignment work: mapping which alignment
failures are interface-recoverable and which require deeper
intervention. This thesis offers one instance of that mapping,
applied to a domain where the answer turned out to be interface.

%% -----------------------------------------------------------------------
\section{Closing}
\label{sec:conc_closing}

\Cref{chap:introduction} opened with a puzzle: large language models
are deployed as personal-decision advisors despite an entropy gap that
compresses roughly 50/50 community judgements into near-100/0 LLM
recommendations.

The resolution offered by this thesis is structural rather than
corrective. The LLMs are not wrong about pluralism. When asked for a
distribution, they produce one that closely matches the community
baseline on the same posts. They are deployed within an interface
that does not ask. Single-answer queries compress the pluralism the
model would otherwise express; the fix, for this class of alignment
failure, is not a better model but a better query.

When deployed as advisors, LLMs do not need to retrain in order to
express the pluralism their outputs can represent --- they need to be
asked for it.
