%% Thesis abstract (front matter). No internal-state claims: all results
%% are scoped to observable output behaviour. Numbers are the locked
%% figures from CLAUDE.md / data/.
\begin{abstract}
Large language models are increasingly deployed as advisors on contested
personal decisions, yet they tend to collapse to a single recommendation
precisely where human communities are genuinely divided. This thesis asks
whether this pluralistic gap is driven by the content of the decision, by
the role the model is assigned, or by the format in which its answer is
elicited. Because token-level probabilities were never accessed, every
claim is scoped to observable output behaviour rather than to any internal
representation.

We construct a corpus of 620 real Reddit decision posts spanning education,
career, and finance, each submitted after the knowledge cutoff of every
model evaluated, which rules out training contamination by construction.
The upvote-weighted distribution of community comments on each post serves
as the alignment reference. Four post-trained large language models from
four developers are evaluated under a forced binary choice and under two
prompt-level interventions: a change of output format (Verbalized Sampling,
applied to the 132 ambiguous posts on which the community is itself split)
and a change of epistemic role (advisor, oracle, and personal framings,
applied to the full corpus). Differences are tested by paired bootstrap
with Benjamini--Hochberg correction, and candidate content moderators by
mixed-effects regression.

Under forced choice the models are near-deterministic (mean entropy 0.046
nats) where the community is broad (0.621 nats). Eliciting a distribution
recovers 94--101\% of this entropy gap across all four models, whereas the
role-framing manipulation is statistically detectable but roughly 16 times
smaller (a reduction of 0.235 versus 0.015 in absolute alignment gap).
Decision-content features moderate the gap only weakly: reversibility,
resource constraint, and socioeconomic context show no significant effect,
while the one detectable content moderator, time horizon, is small relative
to the output-format intervention. The pluralistic gap is therefore format-dependent and
interface-recoverable: an alignment failure of deployment rather than of
training, addressable by how the question is asked rather than by
retraining. These results characterise alignment with the upvote-weighted
Reddit community used here, not with humanity in general.

\vspace{1em}
\noindent\textbf{Keywords:} pluralistic alignment; large language models;
verbalized sampling; output format; decision advice; distributional
alignment; RLHF.
\end{abstract}
